Avoidance in permutations

### 2014-11-07 project2/paper/paper.tex
Resolved some TODOs and cleaned up some notation

In Section \ref{S3} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used throughout this paper.

@@

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Again we use the generating tree. Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Suppose that $\sigma\in S_n$ avoids $\pi$. Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$. Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and see obtain notice that $\sigma\in S_n$ does not avoid $\pi$, a contradiction. The other direction follows because $\mathcal{R}^2=1$. 
\end{proof}

@@

We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. Before formally stating and proving the result, however, we prove the following useful lemma.
\begin{lemma}
\label{lemma1}
The permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the pattern $312$ are precisely those of the form,
$$\pi_1 \pi_2 i$$
the concatenation of $\pi_1, \pi_2$, and $i$, where $\pi_1$ is a permutation of $\{1,2,\ldots,i-1\}$ that avoids the pattern $312$ and $\pi_2$ is a permutation of $\{i+1,\ldots,k+1\}$ that avoids the pattern $312$.
\end{lemma}\todo{Explained the notation for concatenation}

@@

With this lemma proven, we move on to the proof of our main theorem for this section.

\begin{theorem}
$s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
\end{theorem}\todo{should we define $C_n$ separately?}

@@

Using our lemma, we count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation. Consider $\pi$ a permutation of $\{1,2,\ldots, k, k+1\}$ ending in $i$.Let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively. It is clear from the definition of $A$ and $B$ that $A$ and $B$ are disjoint from each other.

@@

Before proving that there exists a bijective mapping between all permutations in $T_n$ that avoid $312$ (call this set $A$) and all permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.

@@

Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$ below.

@@

Consider a permutation $\sigma$ in $T_{2m}$ that avoids $231$. Consider an arbitrary even element $2i$ in $\sigma$. For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to the left of $2i$ in the permutation $\sigma$. First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed.

Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $a > 2i+1$ is an odd number in $A$, then $(2i+1, a, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.

@@

Combining all these facts together, we see that the number of permutations $\sigma$ in $T_n$ that avoid $231$ must be equal to $2^{(2m-2)/2} = 2^{m-1}$, as desired.

@@

We now claim that for every permutation $\sigma$ in $A_m$, the numbers $2m-1$ and $2m$ must be consecutive in the permutation $\sigma$. We argue that this must be true by contradiction. If this were not the case, then it is easy to see that the permutation $\sigma$ would have a sub-sequence in the relative ordering $213$ -- let $j$ be an element that's between $2m-1$ and $2m$ in $\sigma$. Clearly, $j < 2m -1 < 2m$ (since every element in $\sigma$ other than $2m$ is less than $2m-1$), which means $(2m - 1, j, 2m)$ are in the relative ordering $213$. We conclude that $2m-1$ and $2m$ must be consecutive.

Furthermore, we see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$. Furthermore, we claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).

First, note that $2m-1, 2m$ must be inserted to the right of both $2m-3$ and $2m-2$, because of the way $T_{2m}$ is defined. Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ and $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$. Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.

@@

Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different.

>>>>

In Section \ref{S3} we consider sequences $s_n(\pi)$ when $\pi$ is a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers. In Section \ref{S4} we will consider sequences $s_n(\pi)$ when $\pi$ is a permutation of $\{1,2,3,4\}$. In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements. Before proceeding, however, we formally introduce the notation that will be used through the rest of this paper.

@@

\begin{lemma}[Reversing Lemma]
The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
\end{lemma}
\begin{proof}
Again we use the generating tree. Certainly the statement is true for elements of $S_1$. Suppose it is true for elements of $S_{n-1}$. Suppose that $\sigma\in S_n$ avoids $\pi$. Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$. Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction. The other direction follows because $\mathcal{R}^2=1$. 
\end{proof}

@@

We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. Before formally stating and proving the result, however, we prove the following useful lemma.
\begin{lemma}
\label{lemma1}
The permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the pattern $312$ are precisely those of the form,
$$\pi_1 \pi_2 i$$
the concatenation of $\pi_1, \pi_2$, and $i$, where $\pi_1$ is a permutation of $\{1,2,\ldots,i-1\}$ that avoids the pattern $312$ and $\pi_2$ is a permutation of $\{i+1,\ldots,k+1\}$ that avoids the pattern $312$.
\end{lemma}

@@

Before we move on to the proof of our main theorem for this section, we look at the sequence of Catalan numbers.

\begin{definition}
The Catalan numbers are the sequence of positive integers $C_i$ defined as follows,
$$C_0=1, \, C_{n+1}=\sum_{i=1}^n C_iC_{n-i} \, \text{for} \; n \geq 0$$
\end{definition}

\begin{theorem}
$s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

@@

Using our lemma, we count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation. Consider $\pi$ a permutation of $\{1,2,\ldots, k, k+1\}$ ending in $i$. Let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively. It is clear from the definition of $A$ and $B$ that $A$ and $B$ are disjoint from each other.

@@

Before proving that there exists a bijective mapping between the set of permutations in $T_n$ that avoid $312$ (call this set $A$) and the set of permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.

@@

Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$, and hence equivalent, below.

@@

Consider a permutation $\sigma$ in $T_{2m}$ that avoids $231$. Consider an arbitrary even element $2i$ in $\sigma$. For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to its left in the permutation $\sigma$. First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed)

Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $2j+1 > 2i+1$ is an odd number in $A$, then $(2i+1, 2j+1, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.

@@

Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $2 \cdot 2 \cdot \ldots \cdot 2 = 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.

@@

We now claim that for every permutation $\sigma$ in $A_m$, the numbers $2m-1$ and $2m$ must be consecutive in the permutation $\sigma$. We argue that this must be true by contradiction. If this were not the case, then it is easy to see that the permutation $\sigma$ would have a sub-sequence in the relative ordering $213$ -- let $j$ be an element that's between $2m-1$ and $2m$ in $\sigma$. Clearly, $j < 2m -1 < 2m$ (since every element in $\sigma$ other than $2m$ is less than $2m-1$), which means $(2m - 1, j, 2m)$ are in the relative ordering $213$. We conclude that $2m-1$ and $2m$ must be consecutive in $\sigma$.

Furthermore, we see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$. We claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).

First, note that $2m-1, 2m$ must be inserted to the right of both $2m-3$ and $2m-2$, because of the way $T_{2m}$ is defined. Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ or $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$. Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.

@@

Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.
